Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-06-18 MIMOSA.tex
Update figures with new data and start rewrite of results
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\begin{abstract}
Immunological endpoints in vaccine trials are measured through a variety of assays that  provide single--cell measurements of multiple genes and proteins in specific immunological cell populations. %intracellular or cell surface proteins, or mRNA expression levels of genes in single cells from specific immunological cell populations. 
Using single--cell data, we consider the problem of identifying subjects where these cell populations exhibit differential responses under different experimental conditions.  
%While these measurements are usually continuous, they are often discretized and analyzed as counts of single cells. 
For example, in the intracellular cytokine staining assay from flow cytometry, individual cells are classified as either positive or negative for a marker based on a predetermined threshold.  %One such assay is the intracellular cytokine staining assay that
The assay is used to assess an individual's immune response to a vaccine by measuring the number of antigen--specific cells  producing different cytokines in different T--cell subpopulations in response to different antigen stimulations.  Individuals whose T--cells exhibit increased production of a cytokine in response to stimulation are termed ``positive'' for that cytokine, and multiple such ``positivity calls'' are used to identify vaccine responders. 
%Cells are classified as cytokine positive or negative and the resulting The cell counts are used to identify experimental conditions leading to an increase in cytokine producing cells. The ICS assay is generally used to identify ``vaccine responders''. These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline. The rarity of these cell populations makes maximizing the sensitivity and specificity of the assay a primary concern. The standard approach to analyze such data, Fisher's exact test, is problematic for two reasons. It can be overly conservative in detecting true differences for small counts, and the data generally do not meet the assumptions of the test. Specifically, the total cell counts across conditions are generally not fixed since these are generated by independent experiments.  
Here we present a framework based on mixtures of Beta--binomial or Dirichlet--Multinomial distributions for analyzing count data derived from such single--cell assays.  Our method models cellular responses in a marker--specific manner, treating the responding and non--responding observations as separate components in the model. Cell counts from the different experimental conditions are modelled independently, while sharing information across responding and non--responding observations through empirical Bayes priors in order to increase the sensitivity and specificity of positivity calls. We compare our method against Fisher's exact test and show how it can be extended to model multivariate (multiple markers) cellular responses. In simulations and in HIV vaccine trial data we find that our method has higher sensitivity and specificity than Fisher's exact test for positivity calls. 
% simultaneously, sharing information across observations using empirical Bayes methods. The model represents the responder and non-responder observations as separate components in the model, stimulated and unstimulated cell counts are modelled independently. Using simulations and real--world vaccine trial data, we show that our model increases the sensitivity and specificity for positivity calls in ICS assays compared to Fisher's exact test. 
\end{abstract}

@@

\subsubsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
Details about the algorithms are given in Supplementary material. We used the method of Raftery and Lewis \citep{Raftery1,Raftery2} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

\section{Results}
The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation. The unconstrained model was applied to Fluidigm single--cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single--cells.

\subsection{ICS}
\subsubsection*{MIMOSA Outperforms Fisher's Exact Test in Real--World Vaccine Trial Data from HVTN054}

We tested our method on ICS data from HVTN054, a safety and immunogenicity vaccine trial for a replication defective Adenovirus vector HIV vaccine. The data set consisted of 48 individuals who received two doses of vaccine or placebo, and ICS time points for these individuals were available at day 0 and day 28 after vaccination. Responders for each cytokine and stimulation combination were classified using Fisher's exact test and the constrained MIMOSA model at the 1\% FDR level, then the response rates for \textit{any cytokine} being positive within the control and treatment groups were calculated for both methods. Table~\ref{tab:positivityrates} tabulates the response rates for a subset of the ICS data (Gag1, Gag2, Env1, Env2, and Pol2 stimulations, for any cytokine positive, at days 0 and 28, in both CD4 and CD8 T--cell subpopulations).

% latex table generated in R 2.16.0 by xtable 1.7-0 package
% Wed Apr  4 09:53:11 2012
\begin{table}[ht]
\begin{center}
\caption{Response rates for Pol2, Gag1, Gag2, Env1, and Env2 stimulated CD4+ and CD8+ T--cells on Day 0 and Day 28. C, T1, T2 are the Control, and increasing dose treatment groups, respectively. No response is expected on Day 0 in any group, while a response is expected on Day 28 in the treatment groups.}\label{tab:positivityrates}
\begin{tabular}{rrrrrrrrr}
  \hline
 &&\multicolumn{3}{c}{Day 0}&\multicolumn{3}{c}{Day 28}\\
  \cmidrule{3-8}
&Model & C  & T1 & T2  &   C  & T1  & T2  \\ 
  \cmidrule{2-8}
\multirow{2}{*}{CD8+ Pol2}&Beta--binomial & 0.00 & \textbf{0.00} & 0.00 &   0.00 & \textbf{0.89} & 0.57 \\ 
& Fisher & 0.00 & 0.05 & 0.00 &   0.00 & 0.83 & 0.57 \\ 
   \hline
\multirow{2}{*}{CD4+ Pol2}&Beta-binomial & 0.00 & \textbf{0.00} & 0.00 &   0.00 & \textbf{0.89} & \textbf{0.71} \\ 
&Fisher & 0.00 & 0.05 & 0.00  & 0.00 & 0.72 & 0.50 \\ 
 \hline
\multirow{2}{*}{CD8+ Gag2}& Beta-binomial & 0.00 & 0.00 & 0.00  & 0.00 & 0.39 & \textbf{0.21} \\ 
& Fisher & 0.00 & 0.00 & 0.00  & 0.00 & 0.39 & 0.14 \\ 
   \hline
\multirow{2}{*}{CD4+ Gag2}& Beta-binomial & 0.00 & \textbf{0.00} & 0.00 &  0.00 & \textbf{0.44} & \textbf{0.50} \\ 
 & Fisher & 0.00 & 0.05 & 0.00 &  0.00 & 0.33 & 0.43 \\ 
   \hline
\multirow{2}{*}{ CD8+ Env1}& Beta-binomial  & 0.00 & 0.00 & 0.00 & 0.29 & \textbf{0.71} & 0.86 \\ 
   & Fisher & 0.00 & 0.00 & 0.00 &  \textbf{0.00} & 0.65 & 0.86 \\ 
   \hline
\multirow{2}{*}{  CD4+ Env1}& Beta-binomial   & 0.00 & 0.00 & 0.00 &  0.14 & \textbf{1.00} & \textbf{0.86} \\ 
 & Fisher & 0.00 & 0.00 & 0.00 &  0.14 & 0.94 & 0.79 \\ 
   \hline
\multirow{2}{*}{CD8+ Env2}  & Beta-binomial & 0.00 & 0.05 & 0.08  & 0.29 & 0.83 & \textbf{0.57} \\ 
  & Fisher & 0.00 & \textbf{0.00} & 0.08 & \textbf{0.14} & 0.83 & 0.43 \\ 
   \hline
\multirow{2}{*}{CD4+ Env2}& Beta-binomial & 0.00 & \textbf{0.00} & 0.00 &  0.00 & \textbf{0.78} & 0.64 \\ 
  & Fisher & 0.00 & 0.05 & 0.00 &  0.00 & 0.72 & 0.64 \\ 
   \hline
\multirow{2}{*}{CD8+ Gag1}& Beta-binomial & 0.00 & \textbf{0.00} & 0.00  & 0.00 & \textbf{0.82} & 0.79 \\ 
 & Fisher & 0.00 & 0.05 & 0.00  & 0.00 & 0.76 & 0.79 \\ 
   \hline
\multirow{2}{*}{CD4+ Gag1}& Beta-binomial & 0.00 & 0.00 & 0.00  & 0.00 & 0.71 & 0.79 \\ 
 & Fisher & 0.00 & 0.00 & 0.00  & 0.00 & 0.71 & 0.79 \\ 
   \hline

\end{tabular}
\end{center}
\end{table}

On day zero, where no response is expected, response rates from the beta--binomial model were either smaller than (5 cases), or no greater (23 cases) than those from Fisher's exact test, with the exception of CD8+ Env2 stimulated T--cells, where Fisher's exact test (0\% response rate) outperformed the beta--binomial model (5\% response rate). On day 28, where a response is expected in the treatment groups, the beta--binomial model had a higher response rate in 12 cases and was no different from Fisher's exact test in the remaining eight cases. In the control group, the beta--binomial model was equal to Fisher's exact test in eight cases, and performed slightly worse than Fisher's exact test in two cases (Env2 and Env1 stimulated CD8+ cells).  
%On day zero, we found that the response rates from the MIMOSA model were zero, and equal to response rates computed from Fisher's exact test across all treatment and control groups. The same was true for the response rate in the control groups at day 28. In the treatment groups, at day 28, the response rates for the MIMOSA model were  generally higher or equal to those for Fisher's exact test, demonstrating the increased sensitivity and specificity of our method.

%The MIMOSA framework generates predictions for individual samples. To better understand the differences between the predictions from the MIMOSA model and Fisher's exact test, we examined the model fit from Gag2 stimulated CD4+ T--cells at day 28 producing IFNg (Figure~\ref{fig:icsdata}). We see that two individuals are predicted as responders by the MIMOSA model which are not detected by Fisher's exact test. Both individuals have a larger proportion of IFNg producing antigen specific T--cells in antigen stimulated than unstimulated samples, as measure by the MAP or ML estimates of the proportions (Figure~\ref{fig:icsdata}).

\subsection{Study 073}
% latex table generated in R 2.16.0 by xtable 1.7-0 package
% Wed Apr 11 15:01:22 2012
\begin{table}[ht]
\begin{center}
\caption{Cytokine-specific response rates for ENV1 stimulated CD4+ T--cells on visits 2 through 14.1. No response is expected on visit 2 (day 0) for any cytokine.}\label{tab:positivityrates2}
\begin{tabular}{rrrrrrrrrrrr}
  \hline
&Visit & 2 & 5 & 6 & 8 & 9 & 11 & 12 & 13 & 14 & 14.1 \\ 
  \hline
\multirow{2}{*}{IL2}&BB & 0.00 & 0.17 & 0.25 & 0.00 & 0.55 & 0.11 & 0.32 & 0.17 & 0.11 & 0.53 \\ 
 &Fisher & 0.01 & 0.17 & 0.25 & 0.00 & 0.55 & 0.15 & 0.30 & 0.20 & 0.13 & 0.47 \\ 
   \hline
  \hline
&Visit & 2 & 5 & 6 & 8 & 9 & 11 & 12 & 13 & 14 \\ 
  \hline
\multirow{2}{*}{TNF}&BB & 0.05 & 0.00 & 0.21 & 0.00 & 0.57 & 0.12 & 0.39 & 0.44 & 0.25 \\ 
  &Fisher & 0.07 & 0.00 & 0.26 & 0.00 & 0.55 & 0.14 & 0.38 & 0.42 & 0.25 \\ 
   \hline
  \hline
 &Visit& 2 & 5 & 6 & 8 & 9 & 11 & 12 & 13 & 14 & 14.1 \\ 
  \hline
\multirow{2}{*}{IFN}&BB & 0.00 & 0.43 & 0.31 & 0.29 & 0.63 & 0.28 & 0.39 & 0.34 & 0.17 & 0.56 \\ 
  &Fisher & 0.00 & 0.29 & 0.29 & 0.43 & 0.60 & 0.26 & 0.37 & 0.32 & 0.16 & 0.44 \\ 
   \hline
\end{tabular}
\end{center}
\end{table}
\subsection{Fluidigm}

\subsection{Simulation Studies}
We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations. Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from the HVTN054 data set, we simulated data from the constrained model with 500 observations, a response rate of 40\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$. The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test. (Figure~\ref{fig:simulations}). The nominal vs observed false discovery rate was also examined for the models and for Fisher's exact test, as well as the accuracy of the estimated prior distributions.

We found that the both the constrained and unconstrained MIMOSA models out--performed Fisher's exact test with respect to sensitivity and specificity at all values of $N$, and that the false discovery rate observed for the mixture model more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations}). Furthermore, both models gave reasonable estimates of the true hyper--parameters (Figure~\ref{fig:simulations}).

Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly,  we examined whether the unconstrained model could be used to accurately fit data generated from the constrained model. We found that the unconstrained model performed as well as the constrained model when fitting data generated from the constrained model (Figure~\ref{fig:simulations}).

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the MLE estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN054 data set (Figure~\ref{fig:simulations_trunc}). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained model fit to constrained data.

@@

\subsection{Parsimonious multivariate model representation}
In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model. The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$. As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters. Here, we propose alternative model parametrizations that could be used to address this problem by reducing the number of required parameters. A simple solution would be to assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$. As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others. As a more flexible solution, we propose to define $K$ main effect parameters, denoted $\{\theta_k:k=1,\dots,K\}$, for the $K$ markers and represent co-expression parameters as a function of these main effects. So a combination of dregree $l$, with positive markers, $k_1$, $k_2$, \dots, $k_l$ would have a hyperparameter defined as $\tau_l(\theta_{k_1}\theta_{k_2}\cdots\theta_{k_l})^{1/l}$, which is the geometric mean of the corresponding main marker effects scaled by a degree effect, $\tau_l$. Note that the scaling degree effect would be the same for any combinations with the same number of positive markers. The rationale being that as the degree of polyfunctionality increases, the number of cells might decrease (or increase in some instances) and the corresponding parameters should be adjusted by a scaling factor to allow for such variation. The choice of the geometric mean versus an arithmetic mean is to account for the fact that when a particular combination of degree 1 is empty, \textit{i.e.} the corresponding $\theta$ value is close to zero, it is unlikely that other combination involving that maker would be large. Our specification would account for this since the product would be near zero if one of the $\theta$ is near zero.  Finally, the null combination would have a single parameter equal to $\theta_0$.

Assuming that the $\theta$'s and $\tau$'s are stimulation specific, this parametrization would result in $4K+3$ parameters, and would scale linearly with the number of markers providing a good alternative between the fully parametrize model and the model with constant rate for moderate to large values of $K$. In this case, we would recommend the use of an MCMC algorithm. The $\theta$'s would be given the same exponential priors than the unconstrained model, \textit{i.e.} exponentional with mean 100. As for the $\tau's$ we would use an exponential distribution with mean 1\footnote{check with real data to see if that makes sense}.

\subsection{ICS example with two markers}
We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles. As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations. We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T--cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}).

The day 28 stimulations show that the posterior probability of response is generally higher for the multivariate MIMOSA model than for the univariate models, and the day zero stimulation (Figure~\ref{fig:polyfunctionality}) shows that the multivariate model has fewer false positives than the univariate MIMOSA model fit to distinct cytokine combinations. We compared multivariate MIMOSA against Fisher's exact test. Fisher's exact test was used to test whether the number of cells expressing a specific combination of cytokines (i.e TNFa+/IL4-, TNFa+/IL4+, and so forth) was significantly higher in the stimulated samples than in the unstimulated samples.  In this comparison, Fisher's exact test detected only one of the 8 responders detected by the MIMOSA model in the Env3 stimulated, day 28 data, and detected two false positives in the day zero data. Fisher's exact test also failed to identify four of the 12 responders identified by the MIMOSA model in the day 28, Env1 stimulated data.

\section{Discussion}
The variety of single--cell assays being adopted by the immunology community is increasing. Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single--cell count data. Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important. Current approaches rely on asymptotic approximations (t--test, or $\chi^2$ test), empirical or ad--hoc methods (2--fold change), or exact tests (Fisher's exact test) where model assumptions are generally not satisfied, all of which can lead to invalid conclusions about the data~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks}. Most importantly, existing classical methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modelling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.

@@

\todo[inline]{Fix references to non--existent figure}
We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).  For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test. These two observations had a 2.32 and 3.08 fold increase in the proportion of cytokine--positive cells in the stimulated vs unstimulated samples, respectively, but differences in absolute positive cell counts of 2-3 cells, although the negative cell counts for the simulated and unstimulated samples were substantially different (32K vs 82K cells, and 65K vs 145K cells, respectively), resulting in the significant difference in proportions. This is a clear example where sharing information through the empirical Bayes framework, and eliminating the conditioning on fixed margins inherent in Fisher's exact test, lead to improved sensitivity on real data for the MIMOSA model. This increased power to detect vaccine responders is important since it impacts decision--making about which vaccine candidates move forward to later phase clinical trials.

@@

%The multivariate MIMOSA model detected two additional responders than Fisher's exact test at the 1\% FDR level (Table~\ref{tab:MDFisher}, top left). The univariate MIMOSA model applied to each cytokine combination individually also outperformed Fisher's exact test, detecting one more responder for the IL4 and TNFa combination (Table~\ref{tab:MDFisher} bottom right), and two more responders for the TNFa and not IL4 cytokine combination (Table~\ref{tab:MDFisher} bottom left).  In the multivariate case, the two responders detected by MIMOSA that were missed by Fisher's exact test had a  2.4--fold and 2.5--fold increase in the proportion of TNFa+/IL4- cells in the stimulated samples than the unstimulated samples. These are the same two samples detected by the univariate MIMOSA model for TNFa+/IL4- cells, one of which also exhibits a 1.7--fold increase in IL4+/TNFa- expression between stimulated and unstimulated samples that was detected by univariate MIMOSA, but not detected by Fisher's exact test.
%\begin{table}[ht]
%\begin{center}
%\begin{minipage}{0.9\linewidth}
%\caption{Confusion matrices comparing responders and non--responders identified by the MIMOSA models and Fisher's exact test. Top left: Fisher's exact test vs the multivariate MIMOSA model for any combination of cytokines.Top right: Fisher's exact test vs MIMOSA for the univariate combination of IL4 and not TNFa. Bottom left: Fisher's exact test vs univariate MIMOSA for TNFa and not IL4. Bottom right: Fisher's exact test vs univariate MIMOSA for IL4 and TNFa. False discovery rate thresholds were set at 1\% for each cytokine combination (except in the multivariate model, where it was 1\% for each observation). }\label{tab:MDFisher}
%\end{minipage}
%\begin{minipage}{0.49\linewidth}
%\center
%\begin{tabular}{rrr}
%  \hline
%  \multicolumn{3}{c}{Any Combination}\\
%  \hline
%&\multicolumn{2}{c}{MIMOSA}\\

>>>>

\begin{abstract}
Immunological endpoints in vaccine trials are measured through a variety of assays that  provide single--cell measurements of multiple genes and proteins in specific immunological cell populations. %intracellular or cell surface proteins, or mRNA expression levels of genes in single cells from specific immunological cell populations. 
Using single--cell data, we consider the problem of identifying subjects where these cell populations exhibit differential responses under different experimental conditions.  
%While these measurements are usually continuous, they are often discretized and analyzed as counts of single cells. 
For example, in the intracellular cytokine staining assay from flow cytometry, individual cells are classified as either positive or negative for a marker based on a predetermined threshold.  %One such assay is the intracellular cytokine staining assay that
The assay is used to assess an individual's immune response to a vaccine by measuring the number of antigen--specific cells  producing different cytokines in different T--cell subpopulations in response to different antigen stimulations.  Individuals whose T--cells exhibit increased production of a cytokine in response to stimulation are termed ``positive'' for that cytokine, and multiple such ``positivity calls'' are used to identify vaccine responders. 
%Cells are classified as cytokine positive or negative and the resulting The cell counts are used to identify experimental conditions leading to an increase in cytokine producing cells. The ICS assay is generally used to identify ``vaccine responders''. These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline. The rarity of these cell populations makes maximizing the sensitivity and specificity of the assay a primary concern. The standard approach to analyze such data, Fisher's exact test, is problematic for two reasons. It can be overly conservative in detecting true differences for small counts, and the data generally do not meet the assumptions of the test. Specifically, the total cell counts across conditions are generally not fixed since these are generated by independent experiments.  
Here we present a framework based on mixtures of Beta--binomial or Dirichlet--Multinomial distributions for analyzing count data derived from such single--cell assays.  Our method models cellular responses in a marker--specific manner, treating the responding and non--responding observations as separate components in the model. Cell counts from the different experimental conditions are modelled independently, while sharing information across responding and non--responding observations through empirical Bayes priors in order to increase the sensitivity and specificity of positivity calls. We compare our method against Fisher's exact test, a likelihood ratio test, and ranked log fold changes, and show how it can be extended to model multivariate (multiple markers) cellular responses. In simulations and applied to real data sets, we find that our method has higher sensitivity and specificity than Fisher's exact test and alternative methods. 
% simultaneously, sharing information across observations using empirical Bayes methods. The model represents the responder and non-responder observations as separate components in the model, stimulated and unstimulated cell counts are modelled independently. Using simulations and real--world vaccine trial data, we show that our model increases the sensitivity and specificity for positivity calls in ICS assays compared to Fisher's exact test. 
\end{abstract}

@@

\subsubsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
Details about the algorithms are given in Supplementary material. We used the method of Raftery and Lewis \citep{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

\section{Results}
The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation. The unconstrained model was applied to Fluidigm single--cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single--cells. We also performed simulation studies to assess the performance of the constrained and unconstrained models in a univariate and multivariate settings.

\subsection{ICS}
\subsubsection*{MIMOSA Outperforms Competing Methods on Vaccine Trial Data from Study HVTN065}

We tested our method on ICS data from HVTN065,  a trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen, with 98 individuals. We examined time points from day 0 and day 182 (the primary endpoint).We examined the CD4+ T--cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples. An ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the constrained MIMOSA model compared to Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non--responders.

\subsection{Fluidigm}
We applied the MIMOSA model to Fluidigm single--cell gene expression data from CD8+ T--cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples. The unconstrained MIMOSA model was fit separately to for each stimulation, and the posterior probabilities of response for each gene, the posterior differences in the proportion of cells expressing each gene between stimulated and unstimulated conditions, as well as the posterior log ratio of the proportion of cells expressing each gene in stimulated and unstimulated conditions were calculated and heat maps generated (Figure~\ref{fit:fluidigm} A-C).

\subsection{Simulation Studies}
We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations. Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from real data, we simulated data from the constrained model with 500 observations, a response rate of 60\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$. The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change. (Figure~\ref{fig:simulations}). The nominal vs observed false discovery rate was also examined to assess the model fit.

For both the constrained and unconstrained simulations, MIMOSA out--performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, B). The false discovery rate for MIMOSA more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations} C, D ).
%Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the MLE estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Figure~\ref{fig:simulations_trunc}). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained model fit to constrained data.

@@

%\subsection{Parsimonious multivariate model representation}
%In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model. The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$. As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters. Here, we propose alternative model parametrizations that could be used to address this problem by reducing the number of required parameters. A simple solution would be to assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$. As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others. As a more flexible solution, we propose to define $K$ main effect parameters, denoted $\{\theta_k:k=1,\dots,K\}$, for the $K$ markers and represent co-expression parameters as a function of these main effects. So a combination of dregree $l$, with positive markers, $k_1$, $k_2$, \dots, $k_l$ would have a hyperparameter defined as $\tau_l(\theta_{k_1}\theta_{k_2}\cdots\theta_{k_l})^{1/l}$, which is the geometric mean of the corresponding main marker effects scaled by a degree effect, $\tau_l$. Note that the scaling degree effect would be the same for any combinations with the same number of positive markers. The rationale being that as the degree of polyfunctionality increases, the number of cells might decrease (or increase in some instances) and the corresponding parameters should be adjusted by a scaling factor to allow for such variation. The choice of the geometric mean versus an arithmetic mean is to account for the fact that when a particular combination of degree 1 is empty, \textit{i.e.} the corresponding $\theta$ value is close to zero, it is unlikely that other combination involving that maker would be large. Our specification would account for this since the product would be near zero if one of the $\theta$ is near zero.  Finally, the null combination would have a single parameter equal to $\theta_0$.
%Assuming that the $\theta$'s and $\tau$'s are stimulation specific, this parametrization would result in $4K+3$ parameters, and would scale linearly with the number of markers providing a good alternative between the fully parametrize model and the model with constant rate for moderate to large values of $K$. In this case, we would recommend the use of an MCMC algorithm. The $\theta$'s would be given the same exponential priors than the unconstrained model, \textit{i.e.} exponentional with mean 100. As for the $\tau's$ we would use an exponential distribution with mean 1\footnote{check with real data to see if that makes sense}. 
%\subsection{ICS example with two markers}
%We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles. As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations. We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T--cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}). 
%The day 28 stimulations show that the posterior probability of response is generally higher for the multivariate MIMOSA model than for the univariate models, and the day zero stimulation (Figure~\ref{fig:polyfunctionality}) shows that the multivariate model has fewer false positives than the univariate MIMOSA model fit to distinct cytokine combinations. We compared multivariate MIMOSA against Fisher's exact test. Fisher's exact test was used to test whether the number of cells expressing a specific combination of cytokines (i.e TNFa+/IL4-, TNFa+/IL4+, and so forth) was significantly higher in the stimulated samples than in the unstimulated samples.  In this comparison, Fisher's exact test detected only one of the 8 responders detected by the MIMOSA model in the Env3 stimulated, day 28 data, and detected two false positives in the day zero data. Fisher's exact test also failed to identify four of the 12 responders identified by the MIMOSA model in the day 28, Env1 stimulated data.
\section{Discussion}
The variety of single--cell assays being adopted by the immunology community is increasing. Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single--cell count data. Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important. Current approaches rely on asymptotic approximations (t--test, or $\chi^2$ test), empirical or ad--hoc methods (2--fold change), or exact tests (Fisher's exact test) where model assumptions are generally not satisfied, all of which can lead to invalid conclusions about the data~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks}. Most importantly, existing classical methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modelling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.

@@

\todo[inline]{Fix references to non--existent figure}
%We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).  For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test. These two observations had a 2.32 and 3.08 fold increase in the proportion of cytokine--positive cells in the stimulated vs unstimulated samples, respectively, but differences in absolute positive cell counts of 2-3 cells, although the negative cell counts for the simulated and unstimulated samples were substantially different (32K vs 82K cells, and 65K vs 145K cells, respectively), resulting in the significant difference in proportions. This is a clear example where sharing information through the empirical Bayes framework, and eliminating the conditioning on fixed margins inherent in Fisher's exact test, lead to improved sensitivity on real data for the MIMOSA model. This increased power to detect vaccine responders is important since it impacts decision--making about which vaccine candidates move forward to later phase clinical trials.

@@

%The multivariate MIMOSA model detected two additional responders than Fisher's exact test at the 1\% FDR level (Table~\ref{tab:MDFisher}, top left). The univariate MIMOSA model applied to each cytokine combination individually also outperformed Fisher's exact test, detecting one more responder for the IL4 and TNFa combination (Table~\ref{tab:MDFisher} bottom right), and two more responders for the TNFa and not IL4 cytokine combination (Table~\ref{tab:MDFisher} bottom left).  In the multivariate case, the two responders detected by MIMOSA that were missed by Fisher's exact test had a  2.4--fold and 2.5--fold increase in the proportion of TNFa+/IL4- cells in the stimulated samples than the unstimulated samples. These are the same two samples detected by the univariate MIMOSA model for TNFa+/IL4- cells, one of which also exhibits a 1.7--fold increase in IL4+/TNFa- expression between stimulated and unstimulated samples that was detected by univariate MIMOSA, but not detected by Fisher's exact test.
%
